TODO: A Survey of StarCraft AI Techniques

### 2013-02-06 survey.tex
ncorporated DAve's input for the AIIDE competitions

\section{Existing work on RTS Game AI - \color{blue}{Resolutions}}\label{sec:review}

@@

% PLANNING:
Approaches using planning techniques have also been explored in the literature. For example Onta\~{n}\'{o}n et al. \cite{CBR_Planning} explored the use of real-time case-based planning (CBP) in the domain of Wargus (a Warcraft II clone). In their work, they used human demonstration to learn plans, that are then composed at run-time in order to form full-fledges strategies to play the game. In \cite{PlanRetrieval} they improve over  their previous CBP approach by using situation assessment for improving the quality and speed of plan retrieval. Hierarchical Task-Network (HTN) planning has also been explored with some success in the context of simpler first-person shooter games. Planning approaches offer more adaptivity of the AI strategy compared to hard-coded approaches. However, the real-time constraints of RTS games limit the planning approaches that can be applied, being HTN and case-based planning the only ones explored so far. Moreover, none of these approaches addresses any timing or scheduling issues, which are key in RTS games.

@@

% GAME TREE SEARCH:
Game tree search techniques have also been explored for tactical decision making. \cite{Chung05} applied Monte-Carlo planning to a capture-the-flag mod of Open RTS. Balla and Fern \cite{UCT} applied upper confidence bounds on trees (UCT: a MCTS algorithm) to tactical assault planning in Wargus. % (Santi: I commented this one, since it's mentioned above) On influence maps, \cite{teamCompositionRTS} studied team composition and maneuvering by learning a self-organizing map, while \cite{HagelbackJ08} presented a multi-agent potential field based bot.

@@

\section{State of the Art Bots for Starcraft - \color{blue}{Information/Productions}}\label{sec:bot}

@@

Figure \ref{fig:bot-architecture} shows some representative examples of the integration architectures used by different bots in the AIIDE and CIG 2011 Starcraft AI competitions \cite{url1,url2}: BroodwarQ \cite{???}, Nova \cite{???}, UAlbertaBot \cite{???}, Skynet \cite{???}, SPAR \cite{???}, AIUR \cite{???}, and BTHAI \cite{???} (see Section \ref{sec:competition} for a comparison of their performance). Each box represents an individual module with a clearly defined task (only modules with a black background can send actions directly to Starcraft). Dashed arrows represent data flow, and solid arrows represent control (when a module can command another module to perform some task). For example, we can see how SPAR is divided in two sets of modules: {\em situation analysis} and {\em decision making}, the first with three modules dedicated to analyze the current situation of the game, and the later with 4 modules dedicated to exploit that information to decide what to do. We can see how the decision making aspect of SPAR is organized hierarchically, with the higher-level module ({\em strategic decision}) issuing commands to the next module ({\em tactical decision}), which sends commands to the next module ({\em action implementation}), and so on. Only the lower-level modules can send actions directly to Starcraft.

@@

{\color{violet}

@@

\subsection{AIIDE}
\label{sec:AIIDE}
this is left empty for now

@@

}

>>>>

\section{Existing work on RTS Game AI}\label{sec:review}

@@

% PLANNING:
Approaches using planning techniques have also been explored in the literature. For example Onta\~{n}\'{o}n et al. \cite{CBR_Planning} explored the use of real-time case-based planning (CBP) in the domain of Wargus (a Warcraft II clone). In their work, they used human demonstration to learn plans, that are then composed at run-time in order to form full-fledges strategies to play the game. In \cite{PlanRetrieval} they improve over  their previous CBP approach by using situation assessment for improving the quality and speed of plan retrieval. Hierarchical Task-Network (HTN) planning has also been explored with some success in the context of simpler first-person shooter games. Planning approaches offer more adaptivity of the AI strategy compared to hard-coded approaches. However, the real-time constraints of RTS games limit the planning approaches that can be applied, being HTN and case-based planning the only ones explored so far. Moreover, none of these approaches addresses any timing or scheduling issues, which are key in RTS games. On notable exception is the work of Churchill and Buro \cite{churchill2011build}, who used planning in order to construct its economic build-orders taking into account timing constraints of the different actions.

@@

% GAME TREE SEARCH:
Game tree search techniques have also been explored for tactical decision making. \cite{Chung05} applied Monte-Carlo planning to a capture-the-flag mod of Open RTS. Balla and Fern \cite{UCT} applied the UCT algorithm (a Monte Carlo Tree Search algorithm) to tactical assault planning in Wargus. % (Santi: I commented this one, since it's mentioned above) On influence maps, \cite{teamCompositionRTS} studied team composition and maneuvering by learning a self-organizing map, while \cite{HagelbackJ08} presented a multi-agent potential field based bot.

@@

\section{State of the Art Bots for Starcraft}\label{sec:bot}

@@

Figure \ref{fig:bot-architecture} shows some representative examples of the integration architectures used by different bots in the AIIDE and CIG Starcraft AI competitions (see Section \ref{sec:competition}): BroodwarQ \cite{???}, Nova \cite{???}, UAlbertaBot \cite{???}, Skynet \cite{???}, SPAR \cite{???}, AIUR \cite{???}, and BTHAI \cite{???}. Each box represents an individual module with a clearly defined task (only modules with a black background can send actions directly to Starcraft). Dashed arrows represent data flow, and solid arrows represent control (when a module can command another module to perform some task). For example, we can see how SPAR is divided in two sets of modules: {\em situation analysis} and {\em decision making}, the first with three modules dedicated to analyze the current situation of the game, and the later with 4 modules dedicated to exploit that information to decide what to do. We can see how the decision making aspect of SPAR is organized hierarchically, with the higher-level module ({\em strategic decision}) issuing commands to the next module ({\em tactical decision}), which sends commands to the next module ({\em action implementation}), and so on. Only the lower-level modules can send actions directly to Starcraft.

@@

@@

The following subsections summarize the results of all the Starcraft AI competitions held at the AIIDE (Artificial Intelligence for Interactive Digital Entertainment) and CIG (Computational Intelligence in Games) conferences during the past years.

\subsection{AIIDE}\label{sec:AIIDE}

Started in 2010, the AIIDE StarCraft AI Competition\footnote{\url{http://www.StarCraftAICompetition.com}} is the most
well known and longest running StarCraft AI Competition in the world. Each year
AI bots are submitted by competitors to do battle within the retail version of
StarCraft: BroodWar, with prizes supplied by Blizzard Entertainment. This
competition has been made possible by the BWAPI StarCraft programming interface,
which allows players to write programs which can retrieve game data and issue
game commands to StarCraft via an easy to use C++ API.

The first competition in 2010 was organized and run by Ben Weber in the Expressive
Intelligence Studio at University of California , Santa Cruz\footnote{\url{http://eis.ucsc.edu/StarCraftAICompetition}}. 26 total submissions
were received from around the world. As this was the first year of the competition,
and little infrastructure had been created, each game of the tournament was run 
manually on two laptop computers and monitored by Ben to record the results. Also,
no persistent data was kept for bots to learn about opponents between matches.

The 2010 competition had 4 different tournament categories in which to compete. Tournament 1
was a flat-terrain unit micromanagement battle consisting of four separate unit
composition games. Of the six competitors, FreSCBot won the competition with
Sherbrooke coming in 2nd place. Tournament 2 was another micro-focused game with
non-trivial terrain. Two competitors submitted for this category, with FreSCBot
once again coming in 1st by beating Sherbrooke.

Tournament 3 was a tech-limited StarCraft game on a single known map with no fog of war enforced. Players
were only allowed to choose the Protoss race, with no late game units allowed. 8 bots
faced off in this double-elimination tournament with mimicbot taking first place over botnik in the final.
As this was a perfect information variant of StarCraft, mimicbot adopted a strategy of
"mimics its opponent's build order, gaining an economic advantage whenever possible" which
worked quite well.

Tournament 4 was the complete game of {\em StarCraft: BroodWar} with fog of war enforced. The tournament
was run with a random pairing double-elimination format with each match being best of 5 games.
Competitors could play as any of the three races, with the only limitations in gameplay
being those which were considered 'cheating' in the StarCraft community. A map pool of
5 well-known professional maps was announced to competitors in advance, with a random map being chosen for each game.

Results are shown in Table \ref{tab:aiide2010}. The team that won was Overmind\footnote{\url{http://overmind.cs.berkeley.edu}}, from University of California, Berkeley. Using the Zerg race, their
strategy was to defend early aggression with zergling units while amassing mutalisk units
which they used to contain and eventually defeat their opponents. The mutalisk is a very fast and agile
flying unit which is able to attack while moving with no drawback, which makes them quite a powerful
unit when controlled by a computer, which has super-human input speeds. Overmind used a potential-field
based micromanagement system to guide their mutalisks, which led them to victory. Krasi0 came in 2nd place
with a standard defensive Terran opening strategy which transitioned into mech play in the late game.

\begin{table}[bth]
\caption{Results of the AIIDE 2010 competition.}
\label{tab:aiide2010}
\begin{small}
\begin{center}
% bracket A
\begin{tabular}{|l|c|}
\hline
Position & bot\\ \hline
1 & Overmind \\
2 & Krasi0 \\
3 & Chronos \\ \hline
\end{tabular}
\end{center}
\end{small}
\end{table}

In 2011 the competition was hosted by the University of Alberta, with organization by Michael Buro and
David Churchill\footnote{\url{https://skatgame.net/mburo/sc2011/}}. Due to a lack of entrants in tournament categories 1-3 in the 2010 competition, it was
decided that only the full game category would be played in the 2011 competition. Another important
change in the 2011 competition was the introduction of automated tournament-managing software running
StarCraft games simultaneously on 20 computers, allowing a total of 1170 games to be played in the far less
time than the 108 games of the 2010 competition. This increase in games played also allowed the tournament
to switch to a round-robin format, eliminating the ``luck'' factor of the pairings inherent in bracket
style tournaments. The bot which achieved the highest win percentage over the course of the competition would
be determined the winner. Also decided was that the competition would become open-source, in an effort not
only to prevent possible cheating, but to promote healthy competition in future tournaments by giving
newcomers and easier entry point by basing their design off of previous bots.

In the end, Skynet\footnote{\url{http://code.google.com/p/skynetbot/}} won the competition with its solid Protoss play (results are summarized in Table \ref{tab:aiide2011}). The bot executed one of a small set of
strategies randomly at the start of the match based on the map and the race of the opponent. Skynet would
then amass a medium to large sized army and expand before moving out to attack. Good use of of Dragoon range
and kiting micromanagement allowed it to hold off the early aggression of other bots such as UAlbertaBot,
which came in 2nd.

UAlbertaBot used an early zealot-rush strategy to take advantage of the power of early game Protoss units. 
It would send out the first zealots which were made and immediately attack the enemy base, using a unit
counting heuristic to determine whether or retreat or keep pushing. Of note is that UAlbertaBot used an
online planning algorithm to construct all of its economic build-orders\cite{churchill2011build}, as no hard-coded build orders
were used.

AIUR also chose Protoss, with a strategy which was in between Skynet and UAlbertaBot in terms of attack
timings. Of note in these results was that a rock-paper-scissors effect happened among the top 3 finishers.
Of the 30 rounds, Skynet beat UAlbertaBot 26 times, UAlbertaBot beat AIUR 29 times, and AIUR beat Skynet
19 times. Another notable result is that Overmind did not choose to compete despite winning the 2010
competition. After the competition, many bot programmers (including the Overmind team) realized that 
their 2010 strategy was quite easily defeated by early game rushing strategies, and so they submitted
a Terran bot instead, called Undermind, which finished in 7th.

\begin{table}[bth]
\caption{Results of the AIIDE 2011 competition.}
\label{tab:aiide2011}
\begin{small}
\begin{center}
% bracket A
\begin{tabular}{|l|c|c|}
\hline
Position & bot & win \% \\ \hline
1 & Skynet & 88.9\% \\
2 & UAlbertaBot & 79.4\% \\
3 & AIUR & 70.3\% \\
4 & ItayUndermind & 65.8\% \\
5 & EISBot & 60.6\% \\ \hline
\end{tabular}
\end{center}
\end{small}
\end{table}

After the competition was over, a man vs. machine match was held between the winner (Skynet) and an
ex-professional StarCraft player named Oriol Vinyals. Oriol was a competitor in the 2001 World Cyber
Games StarCraft competition, and though he had been out of practice for a few years was still quite a
good player. The match was arranged to see how well StarCraft AI bots had progressed and to see if
they could actually beat a decent human opponent.

For the best-of-three match, Oriol chose his races randomly and ended up beating Skynet in a 2-0. 
In the first match, Oriol played Zerg vs. Skynet's Protoss on Python, a four player map. Oriol chose
to start with a fast expansion strategy and transition into two base mutalisk production. Skynet chose
to rush with a few early zealots, which was luckily the best possible choice given Oriol's strategy.
Skynet's initial attack destroyed Oriol's early defenses, and nearly won the game in the first few
minutes, however it then proceeded to send zealots to attack one at a time rather than group up its
units before moving in, which allowed Oriol to catch up. Once Oriol produced his mutalisks, Skynet did
not produce sufficient air defenses and Oriol quickly destroyed Skynet's base. In the second game,
Oril played Terran, again on Python. After holding off early Dragoon pressure from Skynet, Oriol
moved out with a few marines, medics and tanks. Skynet tried to defend with its army of Dragoons,
however due to poor unit targeting decisions it started to attack useless medics after the marines
had died, rather than the tanks. Oriol overcame the Dragoon army and was victorious. Later analysis
of the match concluded that Skynet, while dominant over the other bots, was unable to properly
adapt and transition into a mid-game strategy in game one once its early pressure failed, and in game
two made a key blunder in unit targeting which cost it the game. The humans were still in command.

The 2012 competition was also hosted by the University of Alberta, with the major difference from the 2011
competition being the addition of persistent storage. Bots could now write information to disk during a
match, and then read the information during other matches, allowing them to adjust strategies based
on previous results. 6 of the 10 entrants used this feature to aid in strategy selection, including the
top 4 finishers. More improvements to the tournament environment also meant that a total of 4240 games
could now be played in the same time period. Results are shown in Table \ref{tab:aiide2012}.

Skynet once again won the competition with its solid Protoss build orders and good Dragoon kiting. AIUR
and UAlbertaBot switched positions from the previous year to come 2nd and 3rd respectively. Both AIUR
and UAlbertaBot used data stored from the results of previous games to select a strategy for future matches.
UAlbertaBot did this using the UCB algorithm, while AIUR used a custom algorithm to achieve similar
results. The effect of this strategy selection process can be seen Figure \ref{fig:aiide2012} which shows bot
win percentages over time. While the earlier rounds of the tournament fluctuated wildly in results,
eventually the results converged to their final values. One of the main reasons for this is due to
the bots learning which strategies to use as the tournament progressed.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/aiide2012.png}
    \caption{Evolution of the win percentage of each bot participating in the AIIDE 2012 competition}
    \label{fig:aiide2012}
\end{figure}

\begin{table}[bth]
\caption{Results of the AIIDE 2012 competition.}
\label{tab:aiide2012}
\begin{small}
\begin{center}
% bracket A
\begin{tabular}{|l|c|c|}
\hline
Position & bot & win \% \\ \hline
1 & Skynet & 84.4\% \\
2 & AIUR & 72.2\% \\
3 & UAlbertaBot & 68.6\% \\
4 & BroodwarBotQ & 59.1\% \\
5 & AdjutantBot & 52.8\% \\ \hline
\end{tabular}
\end{center}
\end{small}
\end{table}

The 2012 man vs. machine match again used the winner of the competition (Skynet), who played against
Mike Lange, aka LRM) Bakuryu. At the time of the match, Bakuryu was an A- ranked Zerg player on ICCup,
and known as one of the best non-Korean Zerg players in the world. Bakuryu was considered much
stronger than Oriol at the time that the match was played, and the results showed that this was true.
In the first game of the best-of-three, Bakuryu made Skynet look quite silly by running around inside
Skynet's base with a small number of zerglings while Skynet's zealots and half of its worked chased
then in vain. After killing off several probes and buying enough time to set up his expansion, he
cleaned up Skynet's army with a flying army of mutalisks. In the second game, Bakuryu contained Skynet
inside its base with a group of zerglings positioned within Skynet's expansion. Skynet then constructed
several Dark Templar and along with some Dragoons and Zealots attacked into Bakuryu's expansion which
was heavily defended, and was crushed almost instantly, allowing Bakuryu's zergling force to finish
off the Protoss base.

In this match it was shown that the true weakness of state of the art StarCraft AI systems was that
humans are very adept at recognizing scripted behaviors and exploiting them to the fullest. A human
player in Skynet's position in the first game would have realized he was being taken advantage of and
adapted his strategy accordingly, however the inability to put the local context (Bakuryu kiting his
units around his base) into the larger context of the game (that this would delay Skynet until
reinforcements arrived) and then the lack of strategy change to fix the situation led to an easy
victory for the human. These problems remain the main challenges in RTS AI today: to both recognize
the strategy and intent of an opponent's actions, and how to effectively adapt your own strategy to 
overcome them.

All results, videos, and replays from the AIIDE StarCraft AI Competition can be found in\footnote{\url{http://www.StarCraftAICompetition.com}}

@@
